Smoking and COVID Living Review (`r paste(current_version)`): a bayesian analysis

### 2021-03-01 reports/manuscript.Rmd
v11

---
title: "Smoking and COVID Living Review (`r paste(current_version)`): a bayesian analysis"
author: "David Simons, Lion Shahab, Jamie Brown and Olga Perski"
date: "`r format(Sys.time(), '%A %d %B %Y')`"
output: 
  html_document:
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: false
    theme: simplex
    css: !expr here::here('css', 'style.css')
    keep_md: false
editor_options: 
  chunk_output_type: inline
bibliography: manuscript.bib
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      warning = F,
                      message = F)
options(width = 180)
unlink('reports/markdown_report_cache', recursive = TRUE)
```

@@

**Findings:** `r paste(current_version)` (searches up to `r paste(tail(search_details$date_screened, 1))`) included `r paste(current_total)` studies with `r paste(nrow(running_meta_count))` ‘good’ and ‘fair’ quality studies included in unadjusted meta-analyses. `r paste(nrow(full_smoking_status))` studies (`r paste(round(nrow(full_smoking_status)/current_total*100,1), "%")`) reported current, former and never smoking status with the remainder using broader categories. Recorded smoking prevalence among people with COVID-19 was generally lower than national prevalence. Current compared with never smokers were at reduced risk of SARS-CoV-2 infection (RR = `r paste(round(current_testing_RR_value, 2))`, 95% Credible Interval (CrI) = `r paste(round(current_testing_lci, 2), round(current_testing_uci, 2), sep = "-")`, τ = `r paste(round(current_testing_lci, 2))`). Data for former smokers were inconclusive (RR = `r paste(round(former_testing_RR_value, 2))`, 95% CrI = `r paste(round(former_testing_lci, 2), round(former_testing_uci, 2), sep = "-")`, τ = `r paste(round(former_testing_lci, 2))`) but favoured there being no important association (`r paste(round(1-m2_a_ecdf(1.1),2), "%", sep = "")` probability of RR ≥1.1). Former compared with never smokers were at increased risk of hospitalisation (RR = `r paste(round(former_hospitalisation_RR_value, 2))`, CrI = `r paste(round(former_hospitalisation_lci, 2), round(former_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(former_hospitalisation_lci, 2))`), greater disease severity (RR = `r paste(round(former_severity_RR_value, 2))`, CrI = `r paste(round(former_severity_lci, 2), round(former_severity_uci, 2), sep = "-")`, τ = `r paste(round(former_severity_lci, 2))`) and mortality (RR = `r paste(round(former_mortality_RR_value, 2))`, CrI = `r paste(round(former_mortality_lci, 2), round(former_mortality_uci, 2), sep = "-")`, τ = `r paste(round(former_mortality_lci, 2))`). Data for current smokers on hospitalisation, disease severity and mortality were inconclusive (RR = `r paste(round(current_hospitalisation_RR_value, 2))`, 95% CrI = `r paste(round(current_hospitalisation_lci, 2), round(current_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(current_hospitalisation_lci, 2))`; RR `r paste(round(current_severity_RR_value, 2))`, 95% CrI = `r paste(round(current_severity_lci, 2), round(current_severity_uci, 2), sep = "-")`, τ = `r paste(round(current_severity_lci, 2))`; RR = `r paste(round(current_mortality_RR_value, 2))`, 95% CrI = `r paste(round(current_mortality_lci, 2), round(current_mortality_uci, 2), sep = "-")`, τ = `r paste(round(current_mortality_lci, 2))`, respectively) but favoured there being no important associations with hospitalisation and mortality (`r paste(round(1-m3_a_ecdf(1.1),2), "%", sep = "")` and `r paste(round(1-m5_a_ecdf(1.1),2), "%", sep = "")` probability of RR ≥1.1, respectively) and a small but important association with disease severity (`r paste(round(1-m7_a_ecdf(1.1),2), "%", sep = "")` probability of RR ≥1.1).

@@

COVID-19 is a respiratory disease caused by the SARS-CoV-2 virus. Large age and gender differences in case severity and mortality have been observed in the ongoing COVID-19 pandemic [@guan_clinical_2020]; however, these differences are currently unexplained. SARS-CoV-2 enters epithelial cells through the angiotensin-converting enzyme 2 (ACE-2) receptor [@hoffmann_sars-cov-2_2020]. Some evidence suggests that gene expression and subsequent receptor levels are elevated in the airway and oral epithelium of current smokers [@brake_smoking_2020; @cai_bulk_2020], which could put smokers at higher risk of contracting SARS-CoV-2. Other studies, however, suggest that nicotine downregulates the ACE-2 receptor [@oakes_nicotine_2018]. These uncertainties notwithstanding, both former and current smoking is known to increase the risk of respiratory viral [@denholm_hospitalised_2010; @abadom_risk_2016] and bacterial [@almirall_proportion_1999; @feldman_cigarette_2013] infections and is associated with worse outcomes once infected. Cigarette smoke reduces the respiratory immune defence through peri-bronchiolar inflammationand fibrosis, impaired mucociliary clearance and disruption of the respiratory epithelium [@dye_occasional_1994]. There is also reason to believe that behavioural factors (e.g. regular hand-to-mouth movements) involved in smoking may increase SARS-CoV-2 infection and transmission in current smokers. However, early data from the COVID-19 pandemic have not provided clear evidence for a negative impact of current or former smoking on SARS-CoV-2 infection or COVID-19 disease outcomes, such as hospitalisation or mortality [@vardavas_covid-19_2020]. It has also been hypothesised that nicotine might protect against a hyper-inflammatory response to SARS-CoV-2 infection, which may lead to adverse outcomes in patients with COVID-19 disease [@farsalinos_editorial_2020].

There are several reviews that fall within the scope of smoking and COVID-19 [@vardavas_covid-19_2020],[@emami_prevalence_2020; @alqahtani_prevalence_2020; @patanavanich_smoking_2020; @berlin_covid-19_2020; @farsalinos_systematic_2020; @grundy_smoking_2020]. We aimed to produce a rapid synthesis of available evidence pertaining to the rates of infection, hospitalisation, disease severity and mortality from SARS-CoV-2/COVID-19 stratified by smoking status. Given the increasing availability of data on this topic, this is a living review with regular updates. As evidence accumulates, the review will be expanded to include studies reporting COVID-19 outcomes by alternative nicotine use (e.g., nicotine replacement therapy or e-cigarettes).

@@

A narrative synthesis was conducted. Data from ‘good’ and ‘fair’ quality studies were pooled in R v.3.6.3 [@team_r_2013]. In a living review where new data are regularly added to the analyses, it may be more appropriate to use a Bayesian(as opposed to frequentist) approach where prior knowledge is used in combination with new data to estimate a posterior risk distribution. A Bayesian approach mitigates the issue of performing multiple statistical tests, which can inflate family-wise error. A series of random-effects hierarchical Bayesian meta-analyses were performed with the brms [@burkner_advanced_2018] package to estimate the relative risk for each comparison with accompanying 95% credible intervals (CrIs). We first defined prior distributions for the true pooled effect size (µ) and the between-study heterogeneity (τ), with µ specified as a normal distribution with a mean equal to the derived point estimate from each comparison of interest in the immediately preceding version of this living review, and τ specified as a half-Cauchy distribution with a mean of 0 and standard deviation of 1. The half-Cauchy distribution was selected to reflect prior knowledge that high levels of between-study heterogeneity are more likely than lower levels. Markov Chain Monte Carlo methods (20,000 burn-ins followed by 80,000 iterations) were then used to generate a risk distribution for each study, in addition to a pooled effect for the posterior risk distribution. We report forest plots with the pooled effect for the posterior risk distribution displayed as the median relative risk with an accompanying 95% CrIs. We used the empirical cumulative distribution function (ECDF) to estimate the probability of there being a 10% reduction or 10% increase in relative risk (RR) (i.e. RR ≥1.1 or RR ≤0.9). Due to a lack of indication as to what constitutes a clinically or epidemiologically meaningful effect (e.g. with regards to onward disease transmission or requirements for intensive care beds), we deemed a 10% change in risk as small but important. Where data were inconclusive (as indicated by CrIs crossing RR = 1.0), to disambiguate whether data favoured no effect or there being a small but important association, we estimated whether there was ≥75% probability of RR ≥1.1 or RR ≤0.9.

@@

Characteristics of included studies are presented in [Table 1](#table-1). Studies were conducted across `r paste(nrow(country))` countries. `r paste(all_countries[[1]][1,2])` studies were conducted in the `r paste(all_countries[[1]][1,1])`, `r paste(all_countries[[2]][1,2])` in `r paste(all_countries[[2]][1,1])`, `r paste(all_countries[[3]][1,2])` in the `r paste(all_countries[[3]][1,1])`, `r paste(all_countries[[4]][1,2])`  in `r paste(all_countries[[4]][1,1])`, `r paste(all_countries[[5]][1,2])`  in `r paste(all_countries[[5]][1,1])`, `r paste(all_countries[[6]][1,2])` in `r paste(all_countries[[6]][1,1])`, `r paste(all_countries[[7]][1,2])` in `r paste(all_countries[[7]][1,1])`, `r paste(all_countries[[8]][1,2])` in `r paste(all_countries[[8]][1,1])`, `r paste(all_countries[[9]][1,2])` in `r knitr::combine_words(unlist(all_countries[[9]][1]))`, `r paste(all_countries[[10]][1,2])` in `r knitr::combine_words(unlist(all_countries[[10]][1]))`, `r paste(all_countries[[11]][1,2])` in `r knitr::combine_words(unlist(all_countries[[11]][1]))`, `r paste(all_countries[[12]][1,2])` in `r knitr::combine_words(unlist(all_countries[[12]][1]))`, with `r paste(all_countries[[13]][1,2])` each from `r knitr::combine_words(unlist(all_countries[[13]][1]))` and a single study from `r paste(nrow(all_countries[[14]][1]))` further countries. The majority of studies used observational designs (see [Supplementary table S1](#supplementary-t1)). `r paste(setting[1,2])` (`r paste(round(setting[1,2]/sum(setting[2])*100, 1))`%) were conducted in `r paste(tolower(setting[1,1]))` settings, `r paste(setting[2,2])` studies (`r paste(round(setting[2,2]/sum(setting[2])*100, 1))`%) included individuals from `r paste(tolower(setting[2,1]))` settings, `r paste(setting[3,2])` studies (`r paste(round(setting[3,2]/sum(setting[2])*100, 1))`%) were conducted exclusively in the `r paste(tolower(setting[3,1]))`, with one study each conducted in a homeless shelter and a quarantine centre, and one study that did not state the study setting. Studies had a median of `r paste(participants[3])` (interquartile range = `r paste(participants[2], participants[5], sep = "-")`) participants. The majority of studies (`r paste(round(sum(diagnostics[5]+diagnostics[6])/sum(diagnostics)*100,1))`%) used reverse transcriptase polymerase chain reaction (RT-PCR) for confirmation of SARS-CoV-2 infection, `r paste(round(sum(diagnostics[1]+diagnostics[6])/sum(diagnostics)*100,1))`% used an antibody test to confirm prior infection and `r paste(round(diagnostics[6]/sum(diagnostics)*100,1))` of studies relied on a combination of RT-PCR or antibody assays.

@@

Categorisation of smoking status was heterogeneous (see [Table 1](#table-1)). `r paste(data_source[1,2])` studies collected data on smoking status through routine electronic health records (EHRs), `r paste(data_source[2,2])` studies used a bespoke case report form for COVID-19 and `r paste(data_source[3,2])` studies did not state the source for information on smoking status. None of the studies verified smoking status biochemically. Notably, only `r paste(nrow(full_smoking_status))` (`r paste(round(nrow(full_smoking_status)/current_total*100,1))`%) studies reported current, former and never smoking status (see [Supplementary table S2a](#supplementary-t2a)), with a further 26 studies reporting only ever and never smoking status (see [Supplementary table S2b](#supplementary-t2b)). The remaining `r paste(nrow(incomplete_smoking_status))` studies reported current, current/former or current and former smoking status but did not explicitly state whether remaining participants were never smokers or if data were missing on smoking status (see [Supplementary table S2c](#supplementary-t2c)). `r paste(nrow(missing_smok))` studies explicitly reported the proportion with missing data on smoking status, which ranged from `r paste(round(minimum,2), "%", sep = "")` to `r paste(round(maximum,2), "%", sep = "")`.

@@

Six studies recorded the use of alternative nicotine products in current and/or former smokers but did not report COVID-19 outcomes stratified by alternative nicotine use [@miyara_low_2020; @rimland_clinical_2020; @girardeau_confirmed_2020; @ebinger_sars-cov-2_2020; @islam_risk_2020; @gallichotte_longitudinal_2020].

@@

Two studies were performed in random, representative population samples and were rated as ‘good’ quality, and `r paste(nrow(fair_quality_number))` studies were rated as ‘fair’ quality, of which `r paste(nrow(running_meta_count))` studies reported results stratified by smoking status for the outcomes of interest and could be included in meta-analyses. The remaining `r paste(nrow(poor_quality_number))` studies were rated as ‘poor’ quality (see [Table 1](#table-1)).

#### [*Table 1.* All studies included in narrative review and meta-analysis]{#table-1}
``` {r table_1}
library(flextable)
library(ftExtra)

tbl_1 <- readRDS(file = here::here("reports", "tables", "table_1_refs.rds"))

flextable(tbl_1) %>%
  colformat_md(j = "ref") %>%
  colformat_double(j = "Sample size")
```

@@

Unadjusted smoking prevalence compared with overall estimates for national adult smoking prevalence split by country and study setting is presented in [Figure 2a](#fig-2a) and [2b](#fig-2b). Lower than expected current smoking prevalence was generally observed, especially in studies with hospitalised samples. Former smoking prevalence was more similar to expected prevalence when reported; however, prevalence was typically higher than national estimates. National smoking prevalence estimates used for comparison are presented in [Supplementary table 3](#supplementary-t3).

@@

Three studies provided data on access to SARS-CoV-2 diagnostic testing for those meeting local testing criteria by smoking status. In a cohort study of US military veterans aged 54-75 [@rentsch_covid-19_2020], current smokers were more likely to receive a test: 42.3% (1,603/3,789) of the sample were current smokers compared with 23.8% of all veterans aged 50+ years using any tobacco product between 2010-201533. In the UK Biobank cohort [@niedzwiedz_ethnic_2020], a multivariable analysis showed former (RR = 1.29, 95% CI = 1.14-1.45, *p* < .001) and current (RR = 1.44, 95% CI = 1.20-1.71, *p* < .001) compared with never smokers to be more likely to receive a test. In an Australian rapid assessment screening clinic for COVID-19 [@trubiano_covid-match65_2020], 9.4% (397/4,226) of the self-referred sample (subsequently assessed by a healthcare professional to decide on testing) were current smokers. Of these self-referrals, healthcare professionals decided that current compared with former or never smokers were less likely to require a test (RR = 0.93, 95% CI = 0.86-1.0, *p* = 0.045).

@@

`r paste(nrow(table_2_word))` studies provided data on SARS-CoV-2 infection for people meeting local testing criteria by smoking status (see [Table 2](#table-2)). Meta-analyses were performed for `r paste(nrow(good_quality_2))` ‘good’ and `r paste(nrow(fair_quality_2))` ‘fair’ quality studies (see [Figure 3](#fig-3) and [4](#fig-4)). Current smokers were at reduced risk of testing positive for SARS-CoV-2 compared with never smokers (RR = `r paste(round(current_testing_RR_value, 2))`, 95% Credible Interval (CrI) = `r paste(round(current_testing_lci, 2), round(current_testing_uci, 2), sep = "-")`, τ = `r paste(round(current_testing_lci, 2))`). The probability of current smokers being at reduced risk of infection compared with never smokers (RR ≤0.9) was `r paste(round(m1_a_ecdf(0.9),2))`%. Former compared with never smokers were at increased risk of testing positive, but data were inconclusive (RR = `r paste(round(former_testing_RR_value, 2))`, 95% CrI = `r paste(round(former_testing_lci, 2), round(former_testing_uci, 2), sep = "-")`, τ = `r paste(round(former_testing_lci, 2))`) and favoured there being no important association. The probability of former smokers being at increased risk of infection (RR ≥1.1) compared with never smokers was (`r paste(round(1-m2_a_ecdf(1.1), 2), "%")` probability of RR ≥1.1). Results were materially unchanged in sensitivity analyses.

####    [*Table 2* SARS-CoV-2 infection by smoking status]{#table-2}
``` {r table_2}
readRDS(file = here::here("reports", "tables", "table_2.rds"))
```

![*Figure 3.* Forest plot for risk of testing positive for SARS-CoV-2 in current vs. never smokers. * Indicates ‘good’ quality studies. The prior from the previous review version (v9) was RR = 0.69.](figure/m1_a.png){#fig-3}

![*Figure 4.* Forest plot for risk of testing positive for SARS-CoV-2 in former vs. never smokers. * Indicates ‘good’ quality studies. The prior from the previous review version (v9) was RR = 1.02.](figure/m2_a.png){#fig-4}

@@

    1.	Infection following viral exposure depends on individual differences in, for example, genetic susceptibility or immunocompetence, which are poorly understood at present. For example, the household secondary attack rate for COVID-19 is estimated at 17% [@fung_household_2020].  

@@

    1.  Smokers with COVID-19 may be less likely to receive a SARS-CoV-2 test or present to hospital due to lack of access to healthcare and may be more likely to die in the community from sudden complications (i.e. self-selection bias) and thus not be recorded [@noauthor_sts_nodate].

>>>>

---
title: 'Smoking and COVID Living Review (`r paste(current_version)`): a bayesian analysis'
author: "David Simons, Lion Shahab, Jamie Brown and Olga Perski"
date: "`r format(Sys.time(), '%A %d %B %Y')`"
output:
  word_document:
    toc: yes
  html_document:
    highlight: zenburn
    toc: yes
    toc_float: yes
    number_sections: no
    theme: simplex
    css: C:/Users/ucbtds4/R Repositories/smoking_covid_bayes/css/style.css
    keep_md: no
editor_options:
  chunk_output_type: inline
bibliography: manuscript.bib
csl: harvard.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      warning = F,
                      message = F)
options(width = 180)
unlink('reports/markdown_report_cache', recursive = TRUE)
```

@@

**Findings:** `r paste(current_version)` (searches up to `r paste(tail(search_details$date_screened, 1))`) included `r paste(current_total)` studies with `r paste(nrow(running_meta_count))` ‘good’ and ‘fair’ quality studies included in unadjusted meta-analyses. `r paste(nrow(full_smoking_status))` studies (`r paste(round(nrow(full_smoking_status)/current_total*100,1), "%", sep = "")`) reported current, former and never smoking status with the remainder using broader categories. Recorded smoking prevalence among people with COVID-19 was generally lower than national prevalence. Current compared with never smokers were at reduced risk of SARS-CoV-2 infection (RR = `r paste(round(current_testing_RR_value, 2))`, 95% Credible Interval (CrI) = `r paste(round(current_testing_lci, 2), round(current_testing_uci, 2), sep = "-")`, τ = `r paste(round(current_testing_tau, 2))`). Data for former smokers were inconclusive (RR = `r paste(round(former_testing_RR_value, 2))`, 95% CrI = `r paste(round(former_testing_lci, 2), round(former_testing_uci, 2), sep = "-")`, τ = `r paste(round(former_testing_tau, 2))`) but favoured there being no important association (`r paste(round(1-m2_a_ecdf(1.1),2)*100, "%", sep = "")` probability of RR ≥1.1). Former compared with never smokers were at increased risk of hospitalisation (RR = `r paste(round(former_hospitalisation_RR_value, 2))`, CrI = `r paste(round(former_hospitalisation_lci, 2), round(former_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(former_hospitalisation_tau, 2))`), greater disease severity (RR = `r paste(round(former_severity_RR_value, 2))`, CrI = `r paste(round(former_severity_lci, 2), round(former_severity_uci, 2), sep = "-")`, τ = `r paste(round(former_severity_tau, 2))`) and mortality (RR = `r paste(round(former_mortality_RR_value, 2))`, CrI = `r paste(round(former_mortality_lci, 2), round(former_mortality_uci, 2), sep = "-")`, τ = `r paste(round(former_mortality_tau, 2))`). Data for current smokers on hospitalisation, disease severity and mortality were inconclusive (RR = `r paste(round(current_hospitalisation_RR_value, 2))`, 95% CrI = `r paste(round(current_hospitalisation_lci, 2), round(current_hospitalisation_uci, 2), sep = "-")`, τ = `r paste(round(current_hospitalisation_tau, 2))`; RR `r paste(round(current_severity_RR_value, 2))`, 95% CrI = `r paste(round(current_severity_lci, 2), round(current_severity_uci, 2), sep = "-")`, τ = `r paste(round(current_severity_tau, 2))`; RR = `r paste(round(current_mortality_RR_value, 2))`, 95% CrI = `r paste(round(current_mortality_lci, 2), round(current_mortality_uci, 2), sep = "-")`, τ = `r paste(round(current_mortality_tau, 2))`, respectively) but favoured there being no important associations with hospitalisation and mortality (`r paste(round(1-m3_a_ecdf(1.1),2)*100, "%", sep = "")` and `r paste(round(1-m7_a_ecdf(1.1),2)*100, "%", sep = "")` probability of RR ≥1.1, respectively) and a small but important association with disease severity (`r paste(round(1-m5_a_ecdf(1.1),2)*100, "%", sep = "")` probability of RR ≥1.1).

@@

COVID-19 is a respiratory disease caused by the SARS-CoV-2 virus. Large age and gender differences in case severity and mortality have been observed in the ongoing COVID-19 pandemic [@guan_clinical_2020]; however, these differences are currently unexplained. SARS-CoV-2 enters epithelial cells through the angiotensin-converting enzyme 2 (ACE-2) receptor [@hoffmann_sars-cov-2_2020]. Some evidence suggests that gene expression and subsequent receptor levels are elevated in the airway and oral epithelium of current smokers [@brake_smoking_2020; @cai_bulk_2020], which could put smokers at higher risk of contracting SARS-CoV-2. Other studies, however, suggest that nicotine downregulates the ACE-2 receptor [@oakes_nicotine_2018]. These uncertainties notwithstanding, both former and current smoking is known to increase the risk of respiratory viral [@denholm_hospitalised_2010; @abadom_risk_2016] and bacterial [@almirall_proportion_1999; @feldman_cigarette_2013] infections and is associated with worse outcomes once infected. Cigarette smoke reduces the respiratory immune defence through peri-bronchiolar inflammation and fibrosis, impaired mucociliary clearance and disruption of the respiratory epithelium [@dye_occasional_1994]. There is also reason to believe that behavioural factors (e.g. regular hand-to-mouth movements) involved in smoking may increase SARS-CoV-2 infection and transmission in current smokers. However, early data from the COVID-19 pandemic have not provided clear evidence for a negative impact of current or former smoking on SARS-CoV-2 infection or COVID-19 disease outcomes, such as hospitalisation or mortality [@vardavas_covid-19_2020]. It has also been hypothesised that nicotine might protect against a hyper-inflammatory response to SARS-CoV-2 infection, which may lead to adverse outcomes in patients with COVID-19 disease [@farsalinos_editorial_2020].

There are several reviews that fall within the scope of smoking and COVID-19 [@vardavas_covid-19_2020; @emami_prevalence_2020; @alqahtani_prevalence_2020; @patanavanich_smoking_2020; @berlin_covid-19_2020; @farsalinos_systematic_2020; @grundy_smoking_2020]. We aimed to produce a rapid synthesis of available evidence pertaining to the rates of infection, hospitalisation, disease severity and mortality from SARS-CoV-2/COVID-19 stratified by smoking status. Given the increasing availability of data on this topic, this is a living review with regular updates. As evidence accumulates, the review will be expanded to include studies reporting COVID-19 outcomes by alternative nicotine use (e.g., nicotine replacement therapy or e-cigarettes).

@@

A narrative synthesis was conducted. Data from ‘good’ and ‘fair’ quality studies were pooled in R v.3.6.3 [@team_r_2013]. In a living review where new data are regularly added to the analyses, it may be more appropriate to use a Bayesian (as opposed to frequentist) approach where prior knowledge is used in combination with new data to estimate a posterior risk distribution. A Bayesian approach mitigates the issue of performing multiple statistical tests, which can inflate family-wise error. A series of random-effects hierarchical Bayesian meta-analyses were performed with the brms [@burkner_advanced_2018] package to estimate the relative risk for each comparison with accompanying 95% credible intervals (CrIs). We first defined prior distributions for the true pooled effect size (µ) and the between-study heterogeneity (τ), with µ specified as a normal distribution with a mean equal to the derived point estimate from each comparison of interest in the immediately preceding version of this living review, and τ specified as a half-Cauchy distribution with a mean of 0 and standard deviation of 1. The half-Cauchy distribution was selected to reflect prior knowledge that high levels of between-study heterogeneity are more likely than lower levels. Markov Chain Monte Carlo methods (20,000 burn-ins followed by 80,000 iterations) were then used to generate a risk distribution for each study, in addition to a pooled effect for the posterior risk distribution. We report forest plots with the pooled effect for the posterior risk distribution displayed as the median relative risk with an accompanying 95% CrIs. We used the empirical cumulative distribution function (ECDF) to estimate the probability of there being a 10% reduction or 10% increase in relative risk (RR) (i.e. RR ≥1.1 or RR ≤0.9). Due to a lack of indication as to what constitutes a clinically or epidemiologically meaningful effect (e.g. with regards to onward disease transmission or requirements for intensive care beds), we deemed a 10% change in risk as small but important. Where data were inconclusive (as indicated by CrIs crossing RR = 1.0), to disambiguate whether data favoured no effect or there being a small but important association, we estimated whether there was ≥75% probability of RR ≥1.1 or RR ≤0.9.

@@

Characteristics of included studies are presented in [Table 1](#table-1). Studies were conducted across `r paste(nrow(country))` countries. `r paste(all_countries[[1]][1,2])` studies were conducted in the `r paste(all_countries[[1]][1,1])`, `r paste(all_countries[[2]][1,2])` in `r paste(all_countries[[2]][1,1])`, `r paste(all_countries[[3]][1,2])` in the `r paste(all_countries[[3]][1,1])`, `r paste(all_countries[[4]][1,2])`  in `r paste(all_countries[[4]][1,1])`, `r paste(all_countries[[5]][1,2])`  in `r paste(all_countries[[5]][1,1])`, `r paste(all_countries[[6]][1,2])` in `r paste(all_countries[[6]][1,1])`, `r paste(all_countries[[7]][1,2])` in `r paste(all_countries[[7]][1,1])`, `r paste(all_countries[[8]][1,2])` in `r paste(all_countries[[8]][1,1])`, `r paste(all_countries[[9]][1,2])` in `r knitr::combine_words(unlist(all_countries[[9]][1]))`, `r paste(all_countries[[10]][1,2])` in `r knitr::combine_words(unlist(all_countries[[10]][1]))`, `r paste(all_countries[[11]][1,2])` in `r knitr::combine_words(unlist(all_countries[[11]][1]))`, `r paste(all_countries[[12]][1,2])` in `r knitr::combine_words(unlist(all_countries[[12]][1]))`, with `r paste(all_countries[[13]][1,2])` each from `r knitr::combine_words(unlist(all_countries[[13]][1]))` and a single study from `r paste(nrow(all_countries[[14]][1]))` further countries. The majority of studies used observational designs (see [Supplementary table S1](#supplementary-t1)). `r paste(setting[1,2])` (`r paste(round(setting[1,2]/sum(setting[2])*100, 1))`%) were conducted in `r paste(tolower(setting[1,1]))` settings, `r paste(setting[2,2])` studies (`r paste(round(setting[2,2]/sum(setting[2])*100, 1))`%) included individuals from `r paste(tolower(setting[2,1]))` settings, `r paste(setting[3,2])` studies (`r paste(round(setting[3,2]/sum(setting[2])*100, 1))`%) were conducted exclusively in the `r paste(tolower(setting[3,1]))`, with one study each conducted in a homeless shelter and a quarantine centre, and one study that did not state the study setting. Studies had a median of `r paste(participants[3])` (interquartile range = `r paste(participants[2], participants[5], sep = "-")`) participants. The majority of studies (`r paste(round(sum(diagnostics[5]+diagnostics[6])/sum(diagnostics)*100,1))`%) used reverse transcriptase polymerase chain reaction (RT-PCR) for confirmation of SARS-CoV-2 infection, `r paste(round(sum(diagnostics[1]+diagnostics[6])/sum(diagnostics)*100,1))`% used an antibody test to confirm prior infection and `r paste(round(diagnostics[6]/sum(diagnostics)*100,1), "%", sep = "")` of studies relied on a combination of RT-PCR or antibody assays.

@@

Categorisation of smoking status was heterogeneous (see [Table 1](#table-1)). `r paste(data_source[1,2])` studies collected data on smoking status through routine electronic health records (EHRs), `r paste(data_source[2,2])` studies used a bespoke case report form for COVID-19 and `r paste(data_source[3,2])` studies did not state the source for information on smoking status. None of the studies verified smoking status biochemically. Notably, only `r paste(nrow(full_smoking_status))` (`r paste(round(nrow(full_smoking_status)/current_total*100,1))`%) studies reported current, former and never smoking status (see [Supplementary table S2a](#supplementary-t2a)), with a further 26 studies reporting only ever and never smoking status (see [Supplementary table S2b](#supplementary-t2b)). The remaining `r paste(nrow(incomplete_smoking_status))` studies reported current, current/former or current and former smoking status but did not explicitly state whether remaining participants were never smokers or if data were missing on smoking status (see [Supplementary table S2c](#supplementary-t2c)). `r paste(nrow(missing_smok))` studies explicitly reported the proportion with missing data on smoking status, which ranged from `r paste(round(minimum,1), "%", sep = "")` to `r paste(round(maximum,1), "%", sep = "")`.

@@

Eight studies recorded the use of alternative nicotine products in current and/or former smokers but did not report COVID-19 outcomes stratified by alternative nicotine use [@miyara_low_2020; @rimland_clinical_2020; @girardeau_confirmed_2020; @ebinger_sars-cov-2_2020; @islam_risk_2020; @gallichotte_longitudinal_2020; @kantele_sars-cov-2_2020; @crooks_predicting_2020]. One additional study grouped together current smokers and vapers [@schubl_risk_2020].

@@

Three studies were performed in random or representative population samples and were rated as ‘good’ quality, and `r paste(nrow(fair_quality_number))` studies were rated as ‘fair’ quality, of which `r paste(nrow(running_meta_count))` studies reported results stratified by smoking status for the outcomes of interest and could be included in meta-analyses. The remaining `r paste(nrow(poor_quality_number))` studies were rated as ‘poor’ quality (see [Table 1](#table-1)).

#### [*Table 1.* All studies included in narrative review and meta-analysis]{#table-1}
``` {r table_1, include=T}
library(flextable)
library(ftExtra)
library(tidyverse)

set_flextable_defaults(
  font.family = "Calibri", 
  font.size = 8,
  font.color = "black",
  digits = 2,
  theme_fun = "theme_booktabs"
  )

tbl_1 <- readRDS(file = here::here("reports", "tables", "table_1_refs.rds")) %>%
  arrange(`Date published`) %>%
  mutate(`Lead author` = recode(`Lead author`, "Schubl" = "Schuble*"))

flextable(tbl_1) %>%
  colformat_md(j = "ref") %>%
  colformat_double(j = "Sample size", digits = 0)
```

@@

Unadjusted smoking prevalence compared with overall estimates for national adult smoking prevalence split by country and study setting is presented in [Figure 2a](#fig-2a) and [2b](#fig-2b). Lower than expected current smoking prevalence was generally observed, especially in studies with hospitalised samples. Former smoking prevalence was more similar to expected prevalence when reported; however, study-based prevalence was typically higher than national estimates. National smoking prevalence estimates used for comparison are presented in [Supplementary table 3](#supplementary-t3).

@@

Four studies provided data on access to SARS-CoV-2 diagnostic testing for those meeting local testing criteria by smoking status. In a cohort study of US military veterans aged 54-75 [@rentsch_covid-19_2020], current smokers were more likely to receive a test: 42.3% (1,603/3,789) of the sample were current smokers compared with 23.8% of all veterans aged 50+ years using any tobacco product between 2010-2015 [@odani_tobacco_2018]. In the UK Biobank cohort [@niedzwiedz_ethnic_2020], a multivariable analysis showed former (RR = 1.29, 95% CI = 1.14-1.45, *p* < .001) and current (RR = 1.44, 95% CI = 1.20-1.71, *p* < .001) compared with never smokers to be more likely to receive a test. In an Australian rapid assessment screening clinic for COVID-19 [@trubiano_covid-match65_2020], 9.4% (397/4,226) of the self-referred sample (subsequently assessed by a healthcare professional to decide on testing) were current smokers. Of these self-referrals, healthcare professionals decided that current compared with former or never smokers were less likely to require a test (RR = 0.93, 95% CI = 0.86-1.0, *p* = 0.045). In a further study using the UK Biobank cohort [@didikoglu_early_2021], current (RR = 1.23, 95% CI = 1.19-1.26, *p* < 0.001) and former smokers (RR = 1.20, 95% CI = 1.18-1.23, *p* < 0.001) were more likely to receive a test compared with never smokers.

@@

`r paste(nrow(table_2_word))` studies provided data on SARS-CoV-2 infection for people meeting local testing criteria by smoking status (see [Table 2](#table-2)). Meta-analyses were performed for `r paste(nrow(good_quality_2))` ‘good’ and `r paste(nrow(fair_quality_2))` ‘fair’ quality studies (see [Figure 3](#fig-3) and [4](#fig-4)). Current smokers were at reduced risk of testing positive for SARS-CoV-2 compared with never smokers (RR = `r paste(round(current_testing_RR_value, 2))`, 95% Credible Interval (CrI) = `r paste(round(current_testing_lci, 2), round(current_testing_uci, 2), sep = "-")`, τ = `r paste(round(current_testing_tau, 2))`). The three good quality studies each reported point estimates less than 1, although the CrI for one of the three studies crossed 1. The probability of current smokers being at reduced risk of infection compared with never smokers (RR ≤0.9) was >99%. Former compared with never smokers were at increased risk of testing positive, but data were inconclusive (RR = `r paste(round(former_testing_RR_value, 2))`, 95% CrI = `r paste(round(former_testing_lci, 2), round(former_testing_uci, 2), sep = "-")`, τ = `r paste(round(former_testing_tau, 2))`) and favoured there being no important association. The probability of former smokers being at increased risk of infection (RR ≥1.1) compared with never smokers was `r paste(round(1-m2_a_ecdf(1.1), 2)*100, "%", sep = "")`. Results were materially unchanged in sensitivity analyses.

####    [*Table 2* SARS-CoV-2 infection by smoking status]{#table-2}
``` {r table_2, include=T}
readRDS(file = here::here("reports", "tables", "table_2.rds"))
```

![*Figure 3.* Forest plot for risk of testing positive for SARS-CoV-2 in current vs. never smokers. * Indicates ‘good’ quality studies. The prior from the previous review version (v10) was RR = 0.69.](figure/m1_a.png){#fig-3}

![*Figure 4.* Forest plot for risk of testing positive for SARS-CoV-2 in former vs. never smokers. * Indicates ‘good’ quality studies. The prior from the previous review version (v10) was RR = 1.02.](figure/m2_a.png){#fig-4}

@@

    1.	Infection following viral exposure depends on individual differences in, for example, genetic susceptibility or immunocompetence, which are poorly understood at present and may be confounded with smoking. For example, the household secondary attack rate for COVID-19 is estimated at 17% [@fung_household_2020].  

@@

    1.  Smokers with COVID-19 may be less likely to receive a SARS-CoV-2 test or present to hospital due to lack of access to healthcare and may be more likely to die in the community from sudden complications (i.e. self-selection bias) and thus not be recorded [@brown_smoking_2020].   
